---
output: github_document
---

## **reReg**

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.4.0-6666ff.svg)](https://cran.r-project.org/)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/reReg)](https://cran.r-project.org/package=reReg)
[![packageversion](https://img.shields.io/badge/Package%20version-1.1.5-orange.svg?style=flat-square)](commits/master)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)
[![Build Status](https://travis-ci.org/user/pkg.svg?branch=master)](https://travis-ci.org/user/pkg)
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

### **reReg** implements a collection of regression models for recurrent event process and failure time.
***

#### Installation

You can install and load **reReg** from CRAN using

```{r cran-installation, eval = FALSE}
install.packages("reReg")
library(reReg)
```

You can install reReg from github with:

```{r gh-installation, eval = FALSE}
## install.packages("devtools")
devtools::install_github("stc04003/reReg")
```


#### References:

     Xu, G., Chiou, S.H., Huang, C.-Y., Wang, M.-C. and Yan, J. (2017).
     Joint Scale-change Models for Recurrent Events and Failure Time.
     *Journal of the American Statistical Association* 112.518:
     796-805.

     Lin, D., Wei, L., Yang, I. and Ying, Z. (2000). Semiparametric
     Regression for the Mean and Rate Functions of Recurrent Events.
     *Journal of the Royal Statistical Society: Series B
     (Methodological)*, **62**: 711 - 730.

     Wang, M.C., Qin, J., and Chiang, C.T. (2001). Analyzing Recurrent
     Event Data with Informative Censoring. *Journal of the American
     Statistical Association* **96**455: 1057-1065.

     Ghosh, D. and D.Y. Lin (2003). Semiparametric Analysis of
     Recurrent Events Data in the Presence of Dependent Censoring.
     *Biometrics*, **59**: 877 - 885.

     Huang, C.Y. and Wang, M.C. (2004). Joint Modeling and Estimation
     for Recurrent Event Processes and Failure Time Data. *Journal of
     the American Statistical Association* **99**(468), 1153-1165.
