---
title: "Visualization of recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Visualization of recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE, collapse = TRUE}
library(knitr)
opts_chunk$set(prompt = TRUE, comment = "")
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
```
In this vignette, we demonstrate how to create a `reSurv` object in __`reReg`__ package.
The `reSurv` object is then used to create event plots and cumulative sample mean (CSM) plots.
We will illustrate the usage of our functions with the `readmission` data from the __`frailtypack`__ 
package [@rondeau2012frailtypack, @gonzalez2005sex]. 
The data contains re-hospitalization times after surgery in patients diagnosed with colorectal cancer.
In this data set, the recurrent event is the readmission and the terminal event is either death or end of study.
See `?readmission` for data details.

```{r load-pkg, message = FALSE}
library(reReg)
packageVersion("reReg")
```

```{r data}
data(readmission, package = "frailtypack")
head(readmission)
readmission <- subset(readmission, !(id %in% c(60, 109, 280)))
attach(readmission)
```

## `reSurv`/`Recur` objects 
**<span style="color:red">The `reSurv()` function is being deprecated in Version 1.2.0.</span>**
A replacement of this function is the `Recur()` function imported from the [`reda`](https://github.com/wenjie2wang/reda) package [@R:reda]. 
We give some background on the `reSurv()` function and a brief introduction to the `Recur()` function here.

The usage of `reSurv()` is similar to the
`Surv()` function in the __`survival`__ package.
There are five arguments in `reSurv()` but not all of them need to be specified.
The arguments of `reSurv()` are as follow
```{r reSurv}
args(reSurv)
```
The five arguments are as follows

* `time1` starting time for the gap time between two successive recurrent events
* `time2` ending time for the gap time between two successive recurrent events
* `id` subject's id
* `event` a vector used as the recurrent event indicator; `event = 0` for non-recurrent times (such as the censoring times)
* `status` a binary vector used as the status indicator for the terminal event; `status = 0` for censored time
* `origin` a numerical vector indicating the time origin of subjects

The arguments of `Recur()` are similar to those of `reSurv()`. THey are listed as follow
```{r Recur}
args(Recur)
```
The six arguments are as follows 

* `time` is a numerical vector representing the time of recurrence event or censoring, or a list with elements named `time1` and `time2` for specifying the follow-up of recurrent events. In the latter case, function `%to%` (or `%2%`) can be used to construct such list.  In addition to numeric values, `Date` and `difftime` are allowed and converted to numeric values. An error will be thrown if this argument is not specified.
* `id` specifies the subject identity. It can be numeric vector, character vector, or a factor vector.  If it is left unspecified, `Recur()` will assume that each row represents a subject.
* `event` is a numeric vector that represents the types of the recurrent events. Logical vector is allowed and converted to numeric vector. Non-positive values are internally converted to zero indicating censoring status.
* `terminal` is a numeric vector that represents the status of the terminal event. Logical vector is allowed and converted to numeric vector.  Non-positive values are internally converted to zero indicating censoring status. If a scalar value is specified, all subjects will have the same status of terminal events at their last recurrent episodes. The length of the specified `terminal` should be equal to the number of subjects, or number of data rows. In the latter case, each subject may have at most one positive entry of `terminal` at the last recurrent episode.
* `origin` a numerical vector indicating the time origin of each subject. If a scalar value is specified, all subjects will have the same origin at the specified value.  The length of the specified `origin` should be equal to the number of subjects, or number of data rows. In the latter case, different subjects may have different origins.  However, one subject must have the same origin. In addition to numeric values, `Date` and `difftime` are also supported and converted to numeric values.
* `check` is a character value specifying how to perform the checks for recurrent event data.  Errors or warnings will be thrown, respectively, if the `check` is specified to be `"hard""’"` (default) or `"soft""’"`.  If `check = "none"` is specified, no data checking procedure will be run.

The main arguments of `Recur()` come directly from `reSurv()`, but `Recur()` comes with a comprehensive check for recurrent event data. 
Moreover, `Recur()` returns a `S4` class object that can be used for functions in both the `reReg` package and the `reda` package. 
Readers are referred to a separate [vignette on `Recur()`](ttps://wenjie-stat.me/reda/articles/reda-recur) for a detailed introduction of `Recur()`.


The `Recur` object can be constructed in the same way as an `reSurv` object.
In the current version (1.1.7), the `reSurv()` function can still be used, but the `reSurv` object will be automatically transformed to the corresponding `Recur` object.
For example: 
```{r reSurv2Recur}
invisible(reSurv(t.stop, id, event, death))
identical(suppressWarnings(reSurv(t.stop, id, event, death)), Recur(t.stop, id, event, death))
```
The usage of `Recur()` is very similar to that of `reSurv()`. 
In the following, we introduced different ways to create an `Recur` object.
When the starting times and the ending times of recurrent events are known, 
one can specify both times in `Recur`. 
```{r Recur-obj, eval = FALSE}
Recur(t.start %to% t.stop, id = id, event = event, status = death)
Recur(time = t.stop, id = id, event = event, status = death)
```
The same `Recur` objects can be achieved without specifying the argument names:
```{r Recur-obj-short, eval = FALSE}
Recur(t.start %to% t.stop, id, event, death)
Recur(t.stop, id, event, death)
```
For each subject, the function `Recur()` prints intervals to represent the duration until the next event (a recurrent event or a terminal event).
```{r Recur-obj-short2, eval = TRUE, output.lines = 1:15}
Recur(t.stop, id, event, death)
```
The example above shows patient id \#1 experienced 2 readmission with a terminal event at `t = 1037` (days).
The `+` at `t = 1037` indicates the terminal time was censored, e.g., this patient did not experience the event of interest (death) at `t = 1037`.
Similarly, patient id \#3 has one readmission and died at `t = 783` (days) as indicated by `*` at `783`.
On the other hand patient id \# 4 has more than 3 readmissions and was censored at `t = 2048` (days).
The readmission intervals was suppressed to prevent printing results wider than the screen allowance.
The number of intervals to be printed can be tuned using the `options` and argument `reda.Recur.maxPrint`. 
For example, 
```{r Recur-obj-short2-2, output.lines = 1:15}
options(reda.Recur.maxPrint = 6)
Recur(t.stop, id, event, death)
```

## Event plots {-}
Event plots are a quick and easy way to glance at recurrent event data.
These can be produced by plotting the `Recur` object with `R`'s generic function `plot`, shown in Figure 1.
```{r plot-Recur, fig.cap="Figure 1: Creating an event plot from a `Recur` object."}
reObj <- Recur(t.stop, id, event, death)
plot(reObj)
```


Common graphical options like `xlab`, `ylab`, `main`, and more can be directly passed down to `plot`. 
```{r plot-Recur2, fig.cap="Figure 2: Creating an event plot from a `Recur` object with custom labels."}
plot(reObj, cex = 1.5, xlab = "Time in days", ylab = "Patients", 
     main = "Event plot for readmission data", 
     terminal.name = "Death", 
     recurrent.name = "Hospital readmission")
```


Separate (stratified) event plots can be produced with the `plotEvents` function, 
which is a more specialized function for event plots.
To demonstrate this, we first detach `readmission`.
```{r detach}
detach(readmission)
```
Unlike the generic plot function, `plotEvents` uses a formula object to specify the stratification.
```{r plotEvents}
args(plotEvents)
```
Here are some examples to re-create Figure 1 with `plotEvents`:
```{r plotEvents-example, eval = FALSE}
plotEvents(reObj)
plotEvents(reObj, data = readmission)
plotEvents(reObj ~ 1, data = readmission)
plotEvents(Recur(t.stop, id, event, death) ~ 1, data = readmission)
```
Figure 2 can be re-created with `plotEvents` in a similar fashion:
```{r plotEvents-example2, eval = FALSE}
plotEvents(Recur(t.stop, id, event, death) ~ 1, data = readmission,
           cex = 1.5, xlab = "Time in days", ylab = "Patients", 
           main = "Event plot for readmission data", 
           terminal.name = "Death", recurrent.name = "Hospital readmission")
```
The `plotEvents` function can also stratify event plots by groups. 
For example, the following can be used to stratify an event plot by `sex`.
```{r plotEvents-gender, fig.cap="Figure 3: Event plot grouped by `sex`"}
plotEvents(Recur(t.stop, id, event, death) ~ sex, data = readmission)
```
Event plot by `sex` and `chemo`:
```{r plotEvents-gender-chemo, fig.cap="Figure 4: Event plot grouped by `sex` and `chemo`."}
plotEvents(Recur(t.stop, id, event, death) ~ sex + chemo, data = readmission)
```



## CSM plots {-}
The non-parametric cumulative sample mean (CSM) function can be used to aid event plots 
to determine visually whether a trend or other pattern exists.
The CSM function is defined as follows:
$$\begin{array}{c}
\hat \mu_n(t) = \int_0^tY_\cdot^{-1}(t)dN_\cdot(t),
\end{array}$$
where $Y_\cdot(t) = \sum_{i=1}^nY_i(t)$ is the total number of subjects who are at risk over $[t, t + dt)$ and
$dN_\cdot(t) = \sum_{i=1}^n dN_i(t)$ is the total number of events over the time interval [0, t].
This estimator is also known as the Nelson-Aalen estimator [@nelson2003recurrent] and 
is also implemented in the `mcf()` function in the **`reda`** package [@R:reda].
Another variation of the CSM function is discussed in [@cook2007statistical] assuming $Y_i(t) = n$ for all $t$.
This variation is equivalent to the Nelson-Aalen estimator without adjusting for risk set. 
We refer the Nelson-Aalen estimator and the Cook-Lawless estimator as the CSM function with (default) and without adjusting for risk set, respectively.

The CSM plot can be created with `R`'s generic function `plot`, shown in Figure 5.
```{r plot-csm, fig.cap = "Figure 5: Creating a CSM plot from a `Recur` object."}
plot(reObj, CSM = TRUE)
```

The CSM plot can be created with the more specialized function, `plotCSM`.
Both `plotEvents` and `plotCSM` return **`ggplot2`** objects.
A side-by-side CSM plot, showing the CSM functions with and without risk adjustment, is shown in Figure 6.
```{r plot-csm2, fig.cap = "Figure 6: Creating a CSM plot from a `Recur` object."}
library(gridExtra)
p1 <- plotCSM(Recur(t.stop, id, event, death) ~ 1, data = readmission, main = "")
p2 <- plotCSM(Recur(t.stop, id, event, death) ~ 1, data = readmission, 
              adjrisk = FALSE, main = "")
grid.arrange(p1, p2, ncol=2)
```
As with the `plotEvents`, `plotCSM` can generate CSM plots given covariates. 
Figure 7 depicts the CSM plots grouped by `sex`.
```{r plot-csm-sex, fig.cap = "Figure 7: CSM plot grouped by `sex`."}
p1 <- plotCSM(Recur(t.stop, id, event, death) ~ sex, data = readmission, main = "")
p2 <- plotCSM(Recur(t.stop, id, event, death) ~ sex, data = readmission, 
              adjrisk = FALSE, main = "")
grid.arrange(p1, p2, ncol=2)
```
As in `plotEvents`, the more complicated 
Figure 8 depicts the CSM plots grouped by `sex` and `chemo`.
```{r plot-csm-sex-chemo, fig.cap = "Figure 8: CSM plots grouped by `sex` and `chemo`."}
p1 <- plotCSM(Recur(t.stop, id, event, death) ~ sex + chemo, data = readmission, main = "")
p2 <- plotCSM(Recur(t.stop, id, event, death) ~ sex + chemo, data = readmission, 
              adjrisk = FALSE, main = "")
grid.arrange(p1, p2, ncol=2)
```

## Multiple recurrent event types
The functions `plotEvents` and `plotCSM` can be used to accommodate recurrent event data with multiple recurrent types.
To illustrate this, we generate hypothetical event types and store these in `event2`.
```{r multi-events, message = FALSE}
set.seed(1)
readmission$event2 <- readmission$event * sample(1:3, 852, TRUE)
```
`plotEvents` and `plotCSM` functions can still be applied when `event` is replaced by `event2`. 
Different recurrent events will be denoted by different colors and shapes.
```{r multi-event-plot, fig.cap = "Figure 9: Event plots with multiple recurrent event types."}
plotEvents(Recur(t.stop, id, event2, death) ~ sex, data = readmission)
```
Default labels can be changed
```{r multi-csm2, fig.cap = "Figure 10: CSM plot with multiple recurrent event types and customized labels."}
plotCSM(Recur(t.stop, id, event2, death) ~ sex, adjrisk = FALSE, data = readmission,
        recurrent.name = "Event types", recurrent.type = c("Type 1", "Type 2", "Type 3"))
```

```{r multi-csm, fig.cap = "Figure 11: CSM plot with multiple recurrent event types."}
library(ggplot2)
p1 <- plotCSM(Recur(t.stop, id, event2, death) ~ sex, data = readmission, main = "") +
  theme(legend.position="none")
p2 <- plotCSM(Recur(t.stop, id, event2, death) ~ sex, data = readmission, 
              adjrisk = FALSE, main = "") +
  theme(legend.position="none")
grid.arrange(p1, p2, ncol = 2)
```



## Reference