---
title: "Regression model for recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Fitting semiparametric regression model to recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "")
```	

The regression model implemented in the __`reReg`__ package accommodate informative censoring 
via the use of a subject-specific frailty variable.
In contrast to existing frailty models, the implemented estimation procedure does not require dis-
tributional information on the frailty variable. Using the borrowing-strength approach in the
estimating procedure, our model allows users to specify any combination of the sub-models
between the recurrent event process and the terminal events when they are fitted jointly.

In this vignette, we demonstrate how to use the `reReg()` function in the __`reReg`__ package
to fit semiparametric regression models to recurrent event data.

## Notations
Let $N(t)$ be the number of recurrent events occurring over the interval $[0, t]$ and
$D$ be the failure time of interest subjects to right censoring by $C$.
Define the composite censoring time $Y = \min(D, C, \tau)$ and
the failure event indicator $\Delta = I\{D\le \min(C, \tau)\}$, where
$\tau$ is the maximum follow-up time.
We assume the recurrent event process $N(\cdot)$ is observed up to $Y$.
Let $X$ be a $p$-dimensional covariate vector.
Consider a random sample of $n$ subjects,
the observed data are independent and identically distributed (iid) copies of
$\{Y_i, \Delta_i, X_i, N_i(t), 0\le t\le Y_i\}$,
where the subscript $i$ denotes the index of a subject for $i= 1, \ldots, n$.
Let $m_i = N_i(Y_i)$ be the number of recurrent events the $i$th subject experienced
before time $Y_i$, then the jump times of $N_i(t)$ give the observed recurrent
event times
$t_{i1}, \ldots, t_{im_i}$ when $m_i > 0$.
Thus, the observed data can also be expressed as iid copies
of $\{Y_i, \Delta_i, X_i, m_i, (t_{i1}, \ldots, t_{m_i})\}$.
The primary interests in recurrent event data analysis is in
making inference about the recurrent event process and the failure event
and understanding the corresponding covariate effects.


## Model assumption
We assume a generalized joint frailty scale-change model for the rate function of the recurrent event process and
the hazard function of the failure time formulated as follow: 
\begin{equation}
  \begin{cases}
      \lambda(t) = \displaystyle Z\lambda_0(te^{X^\top\alpha})e^{X^\top\beta},&\\
      h(t) = \displaystyle Zh_0(te^{X^\top\eta})e^{X^\top\theta}, &t\in[0, \tau],
    \end{cases}
\end{equation}
where $Z$ is a latent shared frailty variable to account for association between the two types of
outcomes, $\lambda_0(\cdot)$ is the baseline rate function,
$h_0(\cdot)$ is the baseline hazard function, and the regression coefficients
$(\alpha, \eta)$ and $(\beta, \theta)$ correspond to the shape and size parameters 
of the rate function and hazard function, respectively.
In contrast to many shared-frailty models that require a parametric assumption,
<!--so that the likelihood method can be used to estimate the joint model,-->
following the idea of @Wang2001,
the `reReg()` function implements semiparametric estimation procedures that do not require the
knowledge about the frailty distribution.
As a result, the dependence between recurrent events and failure event is left unspecified
and the proposed implementations accommodate informative censoring.
The `reReg()` function fits the recurrent event data under the above joint model setting. 
The arguments of `reReg()` are as follows

```{r reReg-args}
library(reReg)
args(reReg)
```

* `formula` a formula object, with the response on the left of a "~" operator, and the predictors on the right. 
The response must be a recurrent event survival object as returned by function `Recur`. 
See the vignette on [Visualization of recurrent event data](https://www.sychiou.com/reReg/articles/reReg-plots.html) or 
[Introduction to formula response function `Recur()`](https://wenjie-stat.me/reda/articles/reda-recur) for examples in creating `Recur` objects.
* `data` an optional data frame in which to interpret the variables occurring in the `formula`.
* `B` a numeric value specifies the number of bootstrap (or resampling) for variance estimation. 
When `B = 0`, variance estimation will not be performed.
* `model` a character string specifying the underlying model.
* `se` a character string specifying the method for standard error estimation. 
* `control` a list of control parameters.

#### Specifying model via the `model` argument
When the interest is the covariate effects on the risk of recurrent events
and the terminal event is treated as nuisances,
`model = "cox"` and `model = "gsc"` give the Cox-type model
and the generalized scale-change rate model considered in
@Wang2001 and @xu2019generalized, respectively.
When the recurrent event process and the terminal events are modeled jointly,
the types of rate function and hazard function can be specified simultaneously
in `model` separated by ```|`''.
The model types for the hazard function are `cox`, `ar`, `am`, and `gsc`.
For examples, the joint frailty Cox-type model of @Huang2004 and the
joint frailty accelerated mean model of @xu2017joint can be called by
`model = "cox|cox"` and `model = "am|am"`, respectively.
Depending on the application, users can specify different model types for the
rate function and the hazard function.
For example, `model = "cox|ar"` postulate a Cox-type proportional model for the
recurrent event rate function and an accelerated rate model for the terminal
event hazard function, or $\alpha = \theta = 0$ in (1).

#### Variance estimation
The nonparametric bootstrap method for clustered data is adopted to estimate the
standard errors of the estimators.
The bootstrap samples are formed by resampling the subjects with replacement
of the same size from the original data.
The above estimating procedures are then applied to each bootstrap sample to
provide one draw of the bootstrap estimate.
With a large number of replicates, the asymptotic variance matrices are
estimated by the sample variance of the bootstrap estimates.
In an attempt to overcome the computational burden in bootstrap variance estimation,
parallel computing techniques based on methods in the `parallel` package will be
applied when `boot.parallel = TRUE}`.
The number of CPU cores used in the parallel computing is controlled by the argument
`boot.parCl}`, whose default value is half of the total number of CPU cores available
on the current host identified by `parallel::detectCores() \%/\% 2}`.
When fitting the generalized scale-change rate model of @xu2019generalized, 
e.g., with `model = "sc"`, 
an efficient resampling-based sandwich estimator is also available via
`se = "sand"`. 
The number of bootstrap or resampling is controlled by the argument `B`.
When `B = 0` (default), the variance estimation procedure will not be performed
and only the point estimates will be returned, which can be useful when
the variance estimation is time consuming.
Future work will be devoted to generalize the efficient resampling-based sandwich estimator to all the sub-models.


#### The `control` list
The complete `control` list consists of the following parameters:
  
  * `eqType` is a character string used to specify whether the log-rank-type (`"logrank"`) 
  or the Gehan-type (`"gehan"`) estimating equation is used in the estimating procedure.
  * `solver`  is a character string used to specify the equation solver used for root search; 
  the available options are `BBsolve`, `dfsane`, `BBoptim`, and `optim` 
  (the first three options loads the corresponding equation solver from package `BB` [@Vara:Gilb:BB:2009]).
  * `tol` is a numerical value used to specify the absolute error tolerance in solving the estimating equations. The default value is $10^{-7}$.
  * `ini` is a list of initial values used in the root-finding algorithms.
The list members `alpha`, `beta`, `eta`, and `theta` correspond to the parameters 
$\alpha$, $\beta$, $\eta$, and $\theta$ in
Model~\eqref{eqn:sc|sc}, respectively. The default values for these initial values are zeros.
  * `boot.parallel` is an logical value indicating whether parallel computation will be applied when `se = "boot"` is called.
  * `boot.parCl` is an integer value specifying the number of CPU cores to be used when `parallel = TRUE`.
  The default value is half the CPU cores on the current host.


The default equation solver (`"BB::dfsane"`) uses
the derivative-free Barzilai-Borwein spectral approach for solving nonlinear equations
implemented in `dfsane()` from the package `BB` [@Vara:Gilb:BB:2009].
Setting `solver = "BB::BBsolve"` calls the wrapper function
`BBsolve()` in the `BB` package to locate a root with different
Barzilai-Borwein steplengths, non-monotonicity parameters, and initialization approaches.
Based on our observation, the `"BB::BBsolve"` algorithm generally exhibited
more reliable convergence but the `solver = "BB::dfsane"` algorithm provides a
better balance between convergence and speed.
The alternative options, `solver = "BB::BBoptim"` and `solver = "optim"`, 
attempts to identify roots by minimizing the $\ell_2$-norm of the estimating functions.
The options `solver = "BB::BBoptim"` and `solver = "optim"` call the `BBoptim()` function from the package `BB`
and the base function `optim()`, respectively.



## Examples
We will illustrate the usage of `reReg` with simulated data generated from the `simGSC()` function. 
Readers are referred to the vignette on [Simulating recurrent event data](https://www.sychiou.com/reReg/articles/reReg-sims.html)
for using `simGSC()` to generate recurrent event data.

#### Joint Cox model of @Huang2004
A simulated data following the joint Cox model can be generated by
```{R, cox, cache = TRUE}
set.seed(1); datCox <- simGSC(200, summary = TRUE)
```
Since the default parameters in `simGSC()` are $\alpha = (0, 0)^\top$, $\beta = (-1, -1)^\top$, $\eta = (0, 0)^\top$, 
and $\theta = (-1, -1)^\top$, the underlying true model has the form:
$$\begin{cases}
\lambda(t) = Z \lambda_0(t)e^{-X_1 - X_2}, &\\
h(t) = Z h_0(t)e^{-X_1 - X_2}, & t\in[0, 60].
\end{cases}$$
The model fit is
```{R, coxfit, cache = TRUE}
fit.cox <- reReg(Recur(t.start %to% t.stop, id, event, status) ~ x1 + x2, 
                 B = 200, data = datCox, model = "cox|cox")
summary(fit.cox)
```
The baseline functions can be plotted via `plot()`: 
```{R, coxplot}
plot(fit.cox)
```

#### Joint accelerated mean model of @xu2017joint
A simulated data following the joint accelerated mean model can be generated by
```{R, am, cache = TRUE}
set.seed(1)
par0 <- list(alpha = c(1, 1), beta = c(1, 1), eta = -c(1, 1), theta = -c(1, 1))
datam <- simGSC(200, par = par0, summary = TRUE)
```
The underlying true model has the form:
$$\begin{cases}
\lambda(t) = Z \lambda_0(te^{X_1 + X_2})e^{X_1 + X_2}, &\\
h(t) = Z h_0(te^{-X_1 - X_2})e^{-X_1 - X_2}, & t\in[0, 60].
\end{cases}$$
The model fit is
```{R, amfit, cache = TRUE}
fit.am <- reReg(Recur(t.start %to% t.stop, id, event, status) ~ x1 + x2,
                B = 200, data = datam, model = "am|am")
summary(fit.am)
```
The baseline functions can be plotted via `plot()`: 
```{R, amplot}
plot(fit.am)
```

#### Joint Cox/accelerated rate model
A simulated model following a special case of the generalized scale-change model 
where the rate function has a Cox structure and the hazard function has an accelerated rate model structure 
can be generated by
```{R, coxar, cache = TRUE}
set.seed(1)
par0 <- list(eta = c(1, 1), theta = c(0, 0))
datCoxAr <- simGSC(200, par = par0, summary = TRUE)
```
The underlying true model has the form:
$$\begin{cases}
\lambda(t) = Z \lambda_0(t)e^{-X_1 - X_2}.&\\
h(t) = Z h_0(te^{X_1 + X_2})& t\in[0, 60].
\end{cases}$$
The model fit is
```{R, coxarfit, cache = TRUE}
fit.CoxAr <- reReg(Recur(t.start %to% t.stop, id, event, status) ~ x1 + x2,
                   B = 200, data = datCoxAr, model = "cox|ar")
summary(fit.CoxAr)
```
The baseline functions can be plotted via `plot()`: 
```{R, coxarplot}
plot(fit.CoxAr)
```

## Other popular methods
Some methods that assumes `Z = 1` and requires independent
censorings are also implemented in `reReg`. 
These includes the methods proposed by 
@lin2000semiparametric, @ghosh2002, and @ghosh2003, that can be called by specifying 
`model = "cox.LWYY"`, `model = "cox.GL"`, and `model = "am.GL"`, respectively.

It is also worth noting that methods in multi-state models could also be applied to analyze recurrent events data.
See the vignette on [Multi-state models and competing risks](https://cran.r-project.org/web/packages/survival/vignettes/compete.pdf)
for more details.
		
## Reference