---
title: "Simulating recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Simulating recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "")
```
In this vignette, we demonstrate how to use the `simSC` function in __`reReg`__ package
to simulate recurrent event data from a scale-change model.
Since the scale-change model includes the Cox-type model and the
accelerated mean model as special cases, `simSC` can also be used to generate data from these submodels.
The `simSC` function allows the censoring time to be non-informative (independent given covariate)
or informative about the recurrent event process. 

## Notations
Suppose recurrent events can potentially be observed in the time period $[0, \tau]$.
For a subject, let $N_i(t)$ be the number of events in interval $[0, t]$,
and $X_i$ is a $p\times 1$ covariate vector.
Let $C_i$ be a non-informative censoring time, which is independent of $N_i(\cdot)$ given $X_i$.
On the contrary, let $D_i$ be a failure time (informative censoring time),
which is associated with $N_i(\cdot)$
even after conditioning on $X$.
Then the follow-up time is defined as $Y = \min(C, D, \tau)$.
The observed data are independent and identically distributed copies of
$\{N_i(t), Y_i, X_i: t\le Y_i, i = 1, \ldots, n\}$.
In the following, we suppress the index for the ease of discussion.
   
## The `simSC` function {-}
The function `simSC` generates the recurrent times from a recurrent event process, $N(t)$,
with a pre-specified rate function, $\lambda(t)$, and the failure time, $D$,
from a pre-specified hazard function, $h(t)$.
By specifying the argument `type` in the function `simSC`,
recurrent times and failure time can be generated from the following models:

 * `type = "cox"` for Cox-type models:
  $$\lambda(t) = Z\lambda_0(t) e^{X^\top\alpha}, h(t) = Zh_0(t) e^{X^\top\beta}.$$
 * `type = "am"` for accelerated mean models:
  $$\lambda(t) = Z\lambda_0(te^{X^\top\alpha})e^{X^\top\alpha}, h(t) = Zh_0(te^{X^\top\beta})e^{X^\top\beta}.$$
 * `type = "sc"` for scale-change models:
  $$\lambda(t) = Z\lambda_0(te^{X^\top\alpha})e^{X^\top\beta}, h(t) = Zh_0(te^{X^\top\beta})e^{X^\top\beta}.$$

The $Z$ is a latent frailty variable.
In `simSC` currently only allows two covariates, i.e., $X = (X_{1}, X_{2})^\top$,
where $X_1$ is a Bernoulli random variable with probability 0.5 and
$X_2$ is a standard normal random variable.
The non-informative censoring time, $C$, is generated separately from an exponential distribution 
with mean 80.
The observed follow-up time is then taken to be $Y = \min(D, C, \tau)$.
We further assume the baseline functions
$$\lambda_0(t) = \frac{2}{1 + t}, h_0(t) = \frac{1}{8(1 + t)}.$$

The arguments in `simSC` are as follow
```{r simSC}
library(reReg)
args(simSC)
```
The arguments are as follows

* `n` number of individual
* `a, b` numeric vectors of parameter of length two.
* `indCen` a logical value indicating whether the censoring assumption is imposed.
  When `indCen = TRUE`, we set $Z = 1$.
  Otherwise, $Z$ is generated from a gamma distribution with mean 1 and variance 0.25
  (e.g., `rgamma(1, 4, 4)`).
* `type` a character string specifying the underlying model.
* `tau` a numeric value specifying the maximum observation time, or $\tau$ in the above notation.
* `summary` a logical value indicating whether a brief data summary will be printed.

## Examples
In the following examples, we simulate recurrent event using `simSC`, with `summary = TRUE`.
```{R set-seed, echo = FALSE}
set.seed(123)
```
Cox-type model:
```{R cox}
dat.cox <- simSC(200, c(-1, 1), c(-1, 1), summary = TRUE)
```
Accelerated mean model:
```{R am}
dat.am <- simSC(200, c(-1, 1), c(-1, 1), type = "am", summary = TRUE)
```
Scale-change model:
```{R sc}
dat.sc <- simSC(200, c(-1, 1), c(-1, 1), type = "sc", summary = TRUE)
```

The output of `simSC` are `tibble` objects.
```{R cox-display}
class(dat.cox)
head(dat.cox)
```
```{R cox-DT}
library(DT)
datatable(dat.cox, options = list(pageLength = 10, scrollX=TRUE)) %>% 
  formatRound(c("Time", "x2"), 3)
```
In this example, subject \#1 experienced 3 recurrent events (at time 0.074, 0.152, and 0.190) and died at time 0.224. 
Similary, subject \#2 experienced 3 recurrent events (at time 1.136, 4.917, and 8.839) and is alive when censored at time 60.