---
title: "Visualization of recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Visualization of recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we demonstrate how to create a *reSurv* object in __*reReg*__ package.
The *reSurv* object is then used to create event plots and cumulative sample mean (CSM) plots.
We will illustrate the usage of our functions with the *readmission* data from the __*frailtypack*__
package [@rondeau2012frailtypack].

```{r data}
library(reReg)
data(readmission, package = "frailtypack")
head(readmission)
```

The usage of is similar to the
*Surv* function in the __*survival*__ package.
There are five arguments in *reSurv*, but not all of them needs to be specified.
The arguments of *reSurv* is as follow
```{r reSurv}
args(reSurv)
```
The five arguments are 

* time1 starting time for the gap time between two successive recurrent events
* time2 ending time for the gap time between two successive recurrent events
* id subject's id
* event a binary vector used as the recurrent event indicator; event = 0 for non-recurrent times (such as the censoring times)
* status a binary vector used as the status indicator for the terminal event; status = 0 for censored time
* origin a numerical vector indicating the time origin of subjects.



## Reference