---
title: "Improving semiparametric estimation of the Cox-type proportional rate model with `reReg`"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Visualization of recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE, collapse = TRUE}
library(knitr)
opts_chunk$set(prompt = TRUE, comment = "")
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
knitr::opts_chunk$set(fig.width = 6, fig.height = 3.5) 
```

The Cox-type proportional rate model (available in `reReg()` via `model = "cox.LWYY"`)
is a common semiparametric regression model for 
recurrent event processes under the noninformative censoring assumption.
Since the construction of the pseudo-partial score function ignores the
dependency among recurrent events, 
the efficiency of the Cox-type proportional rate model can be improved by combining 
a system of weighted pseudo-partial score equations via the generalized method of moments (GMM)
and empirical likelihood (EL) estimation [@huang2022improved].
In this vignette, we demonstrate the proposed GMM and EL procedures in the `reReg` package.
We will illustrate the idea with simulated data generated from the `simGSC()` function.


```{r load-pkg, message = FALSE}
library(reReg)
packageVersion("reReg")
```

## Reference
