---
title: "Visualization of recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Visualization of recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE, collapse = TRUE}
library(knitr)
opts_chunk$set(prompt = TRUE, comment = "")
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
```


In this vignette, we demonstrate how to create event plots and mean cumulative function in __`reReg`__ package.
We will illustrate the usage of our functions with the `readmission` data from the __`frailtypack`__ 
package [@rondeau2012frailtypack, @gonzalez2005sex]. 
The data contains re-hospitalization times after surgery in patients diagnosed with colorectal cancer.
In this data set, the recurrent event is the readmission and the terminal event is either death or end of study.
See `?readmission` for data details.


```{r load-pkg, message = FALSE}
library(reReg)
packageVersion("reReg")
```

```{r data}
data(readmission, package = "frailtypack")
head(readmission)
readmission <- subset(readmission, !(id %in% c(60, 109, 280)))
```

## Event plots

An easy way to glance at recurrent event data is by plotting the event plots.
Event plots can be created by applying the generic function
`plot()` to `Recur` objects directly, as shown in Figure 1.
```{r plot-Recur, fig.cap="Figure 1: Creating an event plot from a `Recur` object.", fig.width = 6, fig.height = 3.5}
reObj <- with(readmission, Recur(t.stop, id, event, death))
plot(reObj)
```
The gray horizontal lines represent each subjects' follow-up times,
the green dots marks the recurrent events, and
the red triangles marks terminal events.
With the default setting, the event plot is arranged so that subjects with longer follow-up times
are presented on the top.
The `plot()` method returns a `ggplot2` object [@R:ggplot] to allow extensive flexibility and customization.
Common graphical options like `xlab`, `ylab`, `main`, and more can be directly passed down to `plot()`. 
```{r plot-Recur2, fig.cap="Figure 2: Creating an event plot from a `Recur` object with custom labels.", fig.width = 6, fig.height = 3.5}
plot(reObj, cex = 1.5, xlab = "Time in days", ylab = "Patients", 
     main = "Event plot for readmission data", 
     terminal.name = "Death", 
     recurrent.name = "Hospital readmission")
```

Applying the `plot()` function to `Recur` objects internally calls the `plotEvent()` function, 
which allows users to configure event plots in a model formula.
The synopsis of `plotEvents()` is:
```{R}
args(plotEvents)
```
The arguments are as follows

  -  `formula` is a formula object, which contains an `Recur` object and categorical covariates separated by `~`, 
  where the categorical covariates are used to stratify the event plot.
  - `data` is an optional data frame in which to interpret the variables occurring in the `formula`.
  - `result` is an optional character string specifying whether the event plot is sorted by the subjects' terminal time.
  The available options are 
      - `increasing` sort the terminal time from in ascending order (default). This places longer terminal times on top.
      - `decreasing` sort the terminal time from in descending order. This places shorter terminal times on top.
      - `none` present the event plots as is, without sorting by the terminal times.
  - `calendarTime` is an optional logical value indicating whether to plot in calendar time. 
  When `calendarTime = FALSE` (default), the event plot will have patient time on the x-axis.
  - `control` is a list of control parameters. 
  - `...` are graphical parameters to be passed to the `plotEvents()` function. 
  
The required argument for `plotEvent()` is `formula`.
The full list of these parameters with their default values are displayed below.
```{R}
args(reReg:::plotEvents.control)
```
The parameters `xlab`, `ylab`, and `main` are adopted from the base
`R` `graphics` package for modifying labels to the x-axis, the y-axis,
and the main title, respectively.
The parameters `terminal.name` and `recurrent.name` are character labels
displayed in the legend.
The parameter `recurrent.type` specifies the recurrent event types to be displayed
in the legend when more than one recurrent event types is specified in the `Recur` object.
The parameter `legend.position` is adopted from the % the `legend.position` components of the
`ggplot2` theme to control the location of the legend box.
The parameters `cex` and `alpha` control the size and the
transparency of the symbols on the event plots, respectively.
The `base_size` parameter controls the base font size in pts.

Stratifying the event plot could provide new insights
into the different covariate effects.
For example, an event plot stratified by the binary variable `sex` can be produced with the following code.
```{r plotEvents-gender, fig.cap="Figure 3: Event plot grouped by `sex`", fig.width = 6, fig.height = 3.5}
plotEvents(Recur(t.stop, id, event, death) ~ sex, data = readmission)
```
Similarly, event plot stratified by `sex` and `chemo` can be produced with the following code.
The additional argument `base_size = 5` is included to display the full legend labels. 
```{r plotEvents-gender-chemo, fig.cap="Figure 4: Event plot grouped by `sex` and `chemo`.", fig.width = 6, fig.height = 3.5}
plotEvents(Recur(t.stop, id, event, death) ~ sex + chemo, data = readmission, base_size = 5)
```

The `plotEvents()` can accommodate multiple recurrent types specified
via the event argument in the `Recur` object. 
To illustrate this feature, we perturb the event indicator so that the value of event 
indicates the index of the different recurrent event. 
The following command is used to create the stratified event plot with different recurrent event.
```{r plotEvents-multiple-events, fig.cap="Figure 5: Event plot with multiple events grouped by `sex`.", fig.width = 6, fig.height = 3.5}
readmission$event2 <- with(readmission, ifelse(t.stop > 500 & event > 0, 2, event))
plotEvents(Recur(t.stop, id, event2, death) ~ sex, data = readmission)
```
The different types of recurrent events are marked in different colors in Figure 5. 
The `plotEvents()` function can also create event plots in calendar time by setting `calendarTime = TRUE`. 
For illustration, we create a new simulated data based on `readmission` with time intervals shift
proportionally by `chemo`. We further convert the time intervals to a `Date` class. 
The construction of the new data and the stratified event plot in Figure 6 is included in the following
```{r plotEvents-calendar, fig.cap="Figure 6: Event plot on calendar time.", fig.width = 6, fig.height = 3.5}
readmission2 <- readmission
readmission2$t.start <- as.Date(readmission2$t.start + as.numeric(readmission2$chemo) * 5, 
  origin = "20-01-01")
readmission2$t.stop <- as.Date(readmission2$t.stop + as.numeric(readmission2$chemo) * 5, 
  origin = "20-01-01")
plotEvents(Recur(t.start %2% t.stop, id, event2, death) ~ sex, data = readmission2, calendarTime = TRUE)
```
In this case, subjects with later censoring times are presented on the top and the date string
are printed on the axis label of the event plot.

## Mean cumulative function  plots 
Let $N(t)$ be the number of recurrent events occurring over the interval $[0, t]$ and
$D$ be the failure time of interest subjects to right censoring by $C$.
Define the composite censoring time $Y = \min(D, C, \tau)$ and
the failure event indicator $\Delta = I\{D\le \min(C, \tau)\}$, where
$\tau$ is the maximum follow-up time.
We assume the recurrent event process $N(\cdot)$ is observed up to $Y$.
Let $X$ be a $p$-dimensional covariate vector.
Consider a random sample of $n$ subjects,
the observed data are independent and identically distributed (iid) copies of
$\{Y_i, \Delta_i, X_i, N_i(t), 0\le t\le Y_i\}$,
where the subscript $i$ denotes the index of a subject for $i= 1, \ldots, n$.
Let $m_i = N_i(Y_i)$ be the number of recurrent events the $i$th subject experienced
before time $Y_i$, then the jump times of $N_i(t)$ give the observed recurrent
event times
$t_{i1}, \ldots, t_{im_i}$ when $m_i > 0$.
Thus, the observed data can also be expressed as iid copies
of $\{Y_i, \Delta_i, X_i, m_i, (t_{i1}, \ldots, t_{m_i})\}$.



Under independence censoring,
a nonparametric estimate of $\Lambda(t)$ known as the Nelson-Aalen estimator
is \citep{lawless1995some}
$$\widehat\Lambda(t) = \sum_{i = 1}^n\int_0^t\frac{d N_i(u)}{\sum_{j = 1}^nI(Y_i \ge u)},$$
which is also known as the mean cumulative function (MCF).
The MCF presents the average number of recurrent events per
subject observed by time~$t$ while adjusting for the risk set.
In the case when all subjects remain at risk of recurrent events
throughout the study, i.e.,  $n = \sum_{j = 1}^nI(Y_i \ge t)$, 
the Nelson-Aalen estimator reduces to the cumulative sample mean function
introduced in Chapter 1 of @cook2007statistical.


The MCF is useful in describing and comparing the average number of events of an individual and between groups.
Thus, it provides additional insights into the longitudinal patterns of the recurrent process.
Under the independent censoring assumption,
the Nelson-Aalen estimate can be created by applying the generic function `plot()` to the `Recur` 
object with an additional argument `mcf = TRUE`.
The following command plots the Nelson-Aalen estimate in Figure 7.
The 95\% confidence interval is enabled by additionally setting
`mcf.conf.int = TRUE`.
```{r plotMCF, fig.cap="Figure 7: Creating a MCF plot from a `Recur` object.", fig.width = 6, fig.height = 3.5}
plot(reObj, mcf = TRUE, mcf.conf.int = TRUE)
```
To create the cumulative sample mean function, one needs to additionally specify
the argument `mcf.adjustRiskset = FALSE`.
Plotting the `Recur` object with the argument `mcf = TRUE` internally
calls the `mcf()` function from the  `reda` package [@R:reda].
The `mcf()` function is imported when the `reReg` package is loaded.
The usage of the `mcf()` function is similar to that of the `plotEvent()`,
where the recurrent event process is specified in a model formula with `Recur`
objects as the formula response and the covariates specified in the model formula
should be a combination of factor variables.
The `mcf()` computes the overall sample MCF when an intercept-only-model is specified.
When covariates are specified in the model formula,
the sample MCF for each level of the combination of the factor variables will be computed.
The following code is used to produce Figure 8,
which displays the Nelson-Aalon estimates for `sex = "Female"`and `sex = "Male"`.
```{r plotMCF-mcf, fig.cap="Figure 8: Creating a MCF plot with `mcf()`", fig.width = 6, fig.height = 3.5}
mcf0 <- mcf(Recur(t.start %2% t.stop, id, event, death) ~ sex, data = readmission)
plot(mcf0, conf.int = TRUE)
```
In the presence of informative censoring, the NPMLE of [@wang1986asymptotic]
can be plotted by first fitting an intercept-only model with `reReg()`,
then applying the generic function `plot()`.
The `Recur` object is used as the formula response.
The following code illustrate this feature with output displayed in Figure 9.
```{r plotMCF-reReg, fig.cap="Figure 8: Creating a MCF plot with `reReg()`", fig.width = 6, fig.height = 3.5}
mcf1 <- reReg(Recur(t.start %2% t.stop, id, event, death) ~ 1, data = readmission)
plot(mcf1)
```
The 95\% confidence interval is computed based on the non-parametric bootstrap
approach with 200 bootstrap replicates as specified by the argument `B = 200`.
Separate MCFs can be created and plotted for each level of a factor variable using
the `subset` option in `reReg()`.
The `basebind()` function can then be applied to combine the MCFs into one plot
to allow easy visual comparison.
The synopsis for `basebind()` is shown below.
```{r plotMCF-basebind}
args(basebind)
```
The arguments are 

  - `...` is a list of `ggplot` objects created by plotting `reReg` objects.
  - `legend.title` is an optional character strings to control the legend title in the combined plot.
  - `legend.labels` is an optional character strings to control the legend labels in the combined plot.
  
When fitting regression models with `reReg()`,
the `baseline()` function can be applied to combine the estimates of baseline functions
across groups.
The following code is used to create Figure~ 9.
where the NPMLEs for `sex = "Female"` and `sex = "Male"` are presented.
The corresponding 95\% confidence intervals are computed with 200 bootstrap replicates.
```{r plotMCF-reReg2, fig.cap="Figure 9: Creating a stratified MCF plot with `reReg()`", fig.width = 6, fig.height = 3.5}
mcf2 <- reReg(Recur(t.start %2% t.stop, id, event, death) ~ 1, subset = sex == "Female", data = readmission, B = 200)
mcf3 <- update(mcf2, subset = sex == "Male")
g1 <- plot(mcf2)
g2 <- plot(mcf3)
basebind(g1, g2, legend.title = "Sex", legend.labels = c("Female", "Male"))
```

## Reference
