---
title: "Visualization of recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Visualization of recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE, collapse = TRUE}
knitr::opts_chunk$set(prompt = TRUE, comment = "")
```
In this vignette, we demonstrate how to create a `reSurv` object in __`reReg`__ package.
The `reSurv` object is then used to create event plots and cumulative sample mean (CSM) plots.
We will illustrate the usage of our functions with the `readmission` data from the __`frailtypack`__ 
package [@rondeau2012frailtypack, @gonzalez2005sex]. 
The data contains re-hospitalization times after surgery in patients diagnosed with colorectal cancer.
In this data set, the recurrent event is the readmission and the terminal event is either death or end of study.
See `?readmission` for data details.

```{r load-pkg, message = FALSE}
library(reReg)
packageVersion("reReg")
```

```{r data}
data(readmission, package = "frailtypack")
head(readmission)
attach(readmission)
```

## `reSurv` objects {-}
The usage of `reSurv` is similar to the
`Surv` function in the __`survival`__ package.
There are five arguments in `reSurv` but not all of them need to be specified.
The arguments of `reSurv` is as follow
```{r reSurv}
args(reSurv)
```
The five arguments are as follows

* `time1` starting time for the gap time between two successive recurrent events
* `time2` ending time for the gap time between two successive recurrent events
* `id` subject's id
* `event` a binary vector used as the recurrent event indicator; `event = 0` for non-recurrent times (such as the censoring times)
* `status` a binary vector used as the status indicator for the terminal event; `status = 0` for censored time
* `origin` a numerical vector indicating the time origin of subjects

There are a few ways to create an `reSurv` object.
When the starting times and the ending times of recurrent events are known, 
one can specify both times in `reSurv`. 
 <!--If the time origins of subjects are equal, as in the `readmission` data, the argument `time2` can be deprecated. --> 
```{r reSurv-obj, eval = FALSE}
reSurv(time1 = t.start, time2 = t.stop, id = id, event = event, status = death)
reSurv(time1 = t.stop, id = id, event = event, status = death)
```
The same `reSurv` objects can be achieved without specifying the argument names:
```{r reSurv-obj-short, eval = FALSE}
reSurv(t.start, t.stop, id, event, death)
reSurv(t.stop, id, event, death)
```
The function `reSurv` prints a list-column `tibble`.
```{r reSurv-obj-short2, eval = TRUE}
reSurv(t.stop, id, event, death)
```
The example above shows patient id \#1 experienced 2 readmission (`tij[1]` is a list of two `double`s) with a terminal event at `t = 1037` (days).
The terminal event was censored (`status = 0`).
Similarly, patient id \#3 has one readmission and died at `t = 783` (days).
On the other hand patient id \# 4 has 4 readmission and was censored at `t = 2048` (days).

## Event plots {-}
Event plots are a quick and easy way to glance at recurrent event data.
These can be produced by plotting the `reSurv` object with `R`'s generic function `plot`, shown in Figure 1.
```{r plot-reSurv, fig.cap="Figure 1: Creating an event plot from a `reSurv` object."}
reObj <- reSurv(t.stop, id, event, death)
plot(reObj)
```


Common graphical options like `xlab`, `ylab`, `main`, and more can be directly passed down to `plot`. 
```{r plot-reSurv2, fig.cap="Figure 2: Creating an event plot from a `reSurv` object with custom labels."}
plot(reObj, cex = 1.5, xlab = "Time in days", ylab = "Patients", 
     main = "Event plot for readmission data", 
     terminal.name = "Death", 
     recurrent.name = "Hospital readmission")
```


Separate (stratified) event plots can be produced with the `plotEvents` function, 
which is a more specialized function for event plots.
To demonstrate this, we first detach `readmission`.
```{r detach}
detach(readmission)
```
Unlike the generic plot function, `plotEvents` uses a formula object to specify the stratification.
```{r plotEvents}
args(plotEvents)
```
Here are some examples to re-create Figure 1 with `plotEvents`:
```{r plotEvents-example, eval = FALSE}
plotEvents(reObj)
plotEvents(reObj, data = readmission)
plotEvents(reObj ~ 1, data = readmission)
plotEvents(reSurv(t.stop, id, event, death) ~ 1, data = readmission)
```
Figure 2 can be re-created with `plotEvents` in a similar fashion:
```{r plotEvents-example2, eval = FALSE}
plotEvents(reSurv(t.stop, id, event, death) ~ 1, data = readmission,
           cex = 1.5, xlab = "Time in days", ylab = "Patients", 
           main = "Event plot for readmission data", 
           terminal.name = "Death", recurrent.name = "Hospital readmission")
```
The `plotEvents` function can also stratify event plots by groups. 
For example, the following can be used to stratify an event plot by `sex`.
```{r plotEvents-gender, fig.cap="Figure 3: Event plot grouped by `sex`"}
plotEvents(reSurv(t.stop, id, event, death) ~ sex, data = readmission)
```
Event plot by `sex` and `chemo`:
```{r plotEvents-gender-chemo, fig.cap="Figure 4: Event plot grouped by `sex` and `chemo`."}
plotEvents(reSurv(t.stop, id, event, death) ~ sex + chemo, data = readmission)
```



## CSM plots {-}
The non-parametric cumulative sample mean (CSM) function can be used to aid event plots 
to determine visually whether a trend or other pattern exists.
The CSM function is defined as follows:
$$\begin{array}{c}
\hat \mu_n(t) = \int_0^tY_\cdot^{-1}(t)dN_i(t),
\end{array}$$
where $Y_\cdot(t) = \sum_{i=1}^nY_i(t)$ is the total number of subjects who are at risk over $[t, t + dt)$ and
$N_i(t)$ is the number of events over the time interval [0, t] for the *i*th process.
This estimator is also known as the Nelson-Aalen estimator [@nelson2003recurrent] and 
is also implemented in the `mcf` function in the **`reda`** package [@R:reda].
Another variation of the CSM function is discussed in [@cook2007statistical] assuming $Y_i(t) = n$ for all $t$.
This variation is equivalent to the Nelson-Aalen estimator without adjusting for risk set. 
We refer the Nelson-Aalen estimator and the Cook-Lawless estimator as the CSM function with (default) and without adjusting for risk set, respectively.

The CSM plot can be created with `R`'s generic function `plot`, shown in Figure 5.
```{r plot-csm, fig.cap = "Figure 5: Creating a CSM plot from a `reSurv` object."}
plot(reObj, CSM = TRUE)
```

The CSM plot can be created with the more specialized function, `plotCSM`.
Both `plotEvents` and `plotCSM` return **`ggplot2`** objects.
A side-by-side CSM plot, showing the CSM functions with and without risk adjustment, is shown in Figure 6.
```{r plot-csm2, fig.cap = "Figure 6: Creating a CSM plot from a `reSurv` object."}
library(gridExtra)
p1 <- plotCSM(reSurv(t.stop, id, event, death) ~ 1, data = readmission, main = "")
p2 <- plotCSM(reSurv(t.stop, id, event, death) ~ 1, data = readmission, 
              adjrisk = FALSE, main = "")
grid.arrange(p1, p2, ncol=2)
```
As with the `plotEvents`, `plotCSM` can generate CSM plots given covariates. 
Figure 7 depicts the CSM plots grouped by `sex`.
```{r plot-csm-sex, fig.cap = "Figure 7: CSM plot grouped by `sex`."}
p1 <- plotCSM(reSurv(t.stop, id, event, death) ~ sex, data = readmission, main = "")
p2 <- plotCSM(reSurv(t.stop, id, event, death) ~ sex, data = readmission, 
              adjrisk = FALSE, main = "")
grid.arrange(p1, p2, ncol=2)
```
As in `plotEvents`, the more complicated 
Figure 8 depicts the CSM plots grouped by `sex` and `chemo`.
```{r plot-csm-sex-chemo, fig.cap = "Figure 8: CSM plots grouped by `sex` and `chemo`."}
p1 <- plotCSM(reSurv(t.stop, id, event, death) ~ sex + chemo, data = readmission, main = "")
p2 <- plotCSM(reSurv(t.stop, id, event, death) ~ sex + chemo, data = readmission, 
              adjrisk = FALSE, main = "")
grid.arrange(p1, p2, ncol=2)
```

## Multiple recurrent event types
The functions `plotEvents` and `plotCSM` can be used to accommodate recurrent event data with multiple recurrent types.
To illustrate this, we generate hypothetical event types and store these in `event2`.
```{r multi-events, message = FALSE}
library(dplyr)
set.seed(1)
readmission <- readmission %>% mutate(event2 = event * sample(1:3, 861, TRUE))
```
`plotEvents` and `plotCSM` functions can still be applied when `event` is replaced by `event2`. 
Different recurrent events will be denoted by different colors and shapes.
```{r multi-event-plot, fig.cap = "Figure 9: Event plots with multiple recurrent event types."}
plotEvents(reSurv(t.stop, id, event2, death) ~ sex, data = readmission)
```
Default labels can be changed
```{r multi-csm2, fig.cap = "Figure 10: CSM plot with multiple recurrent event types and customized labels."}
plotCSM(reSurv(t.stop, id, event2, death) ~ sex, adjrisk = FALSE, data = readmission,
        recurrent.name = "Event types", recurrent.type = c("Type 1", "Type 2", "Type 3"))
```

```{r multi-csm, fig.cap = "Figure 11: CSM plot with multiple recurrent event types."}
library(ggplot2)
p1 <- plotCSM(reSurv(t.stop, id, event2, death) ~ sex, data = readmission, main = "") +
  theme(legend.position="none")
p2 <- plotCSM(reSurv(t.stop, id, event2, death) ~ sex, data = readmission, 
              adjrisk = FALSE, main = "") +
  theme(legend.position="none")
grid.arrange(p1, p2, ncol = 2)
```



## Reference