---
title: "Simulating recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Simulating recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "")
```
In this vignette, we demonstrate how to use the `simSC()` function in __`reReg`__ package
to simulate recurrent event data from a scale-change model.
Since the scale-change model includes the Cox-type model and the
accelerated mean model as special cases, `simSC()` can also be used to generate data from these submodels.
The `simSC()` function allows the censoring time to be non-informative (independent given covariate)
or informative about the recurrent event process. 

## Notations
Suppose recurrent events can potentially be observed in the time period $[0, \tau]$.
For a subject, let $N_i(t)$ be the number of events in interval $[0, t]$,
and $X_i$ is a $p\times 1$ covariate vector.
Let $C_i$ be a non-informative censoring time, which is independent of $N_i(\cdot)$ given $X_i$.
On the contrary, let $D_i$ be a failure time (informative censoring time),
which is associated with $N_i(\cdot)$
even after conditioning on $X$.
Then the follow-up time is defined as $Y = \min(C, D, \tau)$.
The observed data are independent and identically distributed copies of
$\{N_i(t), Y_i, X_i: t\le Y_i, i = 1, \ldots, n\}$.
In the following, we suppress the index for the ease of discussion.
   
## The `simSC` function {-}
The function `simSC` generates the recurrent times from a recurrent event process, $N(t)$,
from a pre-specified rate function, $\lambda(t)$, and the failure time, $D$,
from a pre-specified hazard function, $h(t)$.
A general joint model for the rate function and the hazard function can be formulated in the following:
\begin{equation}
\begin{matrix}
\lambda(t) &= Z\lambda_0(te^{X^\top\alpha})e^{X^\top\beta};\\ 
h(t) &= Zh_0(te^{X^\top\eta})e^{X^\top\theta},
\end{matrix}
\label{eq:joint}
\end{equation}
where $Z$ is a latent shared frailty variable,
$(\alpha, \eta)$ and $(\beta, \theta)$ correspond to the shape and size parameters of the 
rate function and the hazard function, respectively.
The frailty variable $Z$ is used to capture the association between 
the rate function and the hazard function, thus, accommodating informative censoring.
The `simSC()` currently only allows two covariates, i.e., $X = (X_{1}, X_{2})^\top$,
where $X_1$ is a Bernoulli random variable with probability 0.5 and
$X_2$ is a standard normal random variable.
The frailty variable $Z$ is generated from a gamma distribution with mean 1.
The non-informative censoring time, $C$, 
is generated separately from an exponential distribution with mean 80.
The observed follow-up time is then taken to be $Y = \min(D, C, \tau)$.
We further assume the baseline functions
$$\lambda_0(t) = \frac{2}{1 + t}, h_0(t) = \frac{1}{8(1 + t)}.$$

The arguments in `simSC` are as follow
```{r simSC, message = FALSE}
library(reReg)
args(simSC)
```
The arguments are as follows

* `n` is the number of individual
* `a1, a2, b1, b2` are numeric vectors of length two, 
these arguments corresponding to $\alpha$, $\beta$, $\eta$, and $\theta$ from \\eqref{eq:joint}, respectively.
* `type` is a character string specifying the underlying model. 
The rate function type and the hazard function type are separated by a vertical bar `|`, with the rate function on the left. For example, `type = "cox|am"` generates the recurrent process from a Cox model and the terminal event from an accelerated mean model. Setting `type = "cox"` gives `type = "cox|cox"`.
* `zVar` is a numeric value specifying the variance of the frailty variable, $Z$. 
* `tau` is a numeric value specifying the maximum observation time, or $\tau$ in the above notation.
* `summary` is a logical value indicating whether a brief data summary will be printed.

## Examples
In the following examples, we simulate recurrent event data when both the rate function and the hazard function
are Cox-type models and use `summary = TRUE` to display some descriptive statistics about the simulated data.
```{R set-seed}
set.seed(273)
dat.cox <- simSC(200, c(-1, 1), c(-1, 1), summary = TRUE)
```

The output of `simSC()` is a data frame. 
The following table shows that subject \#1 experienced `r sum(subset(dat.cox, id == 1)$event)` recurrent events 
(at times `r round(subset(dat.cox, id == 1)$t.stop[1:sum(subset(dat.cox, id == 1)$event)], 3)`)
and died at time `r round(tail(subset(dat.cox, id == 1)$t.stop, 1), 3)`.
Similarly, subject \#2 experienced `r sum(subset(dat.cox, id == 2)$event)` 
recurrent events (at times `r round(subset(dat.cox, id == 2)$t.stop[1:sum(subset(dat.cox, id == 2)$event)], 3)`)
and is alive when censored at time `r round(tail(subset(dat.cox, id == 2)$t.stop, 1), 3)`.
```{R cox-DT}
library(DT)
datatable(dat.cox, options = list(pageLength = 10, scrollX = TRUE)) %>% 
  formatRound(c("t.stop", "x2"), 3)
```
The event plot for the above simulated data is
```{R cox-event}
plotEvents(Recur(t.stop, id, event, status) ~ 1, data = dat.cox)
```


With the same random seed, the following example generates recurrent event data
when the rate function is an accelerated mean model while the hazard function is a Cox model. 
Compare to the previous example, the difference is in the model structure of the rate function. 
```{R set-seed2}
set.seed(273)
dat.amcox <- simSC(200, c(-1, 1), c(-1, 1), type = "am|cox")
datatable(dat.amcox, options = list(pageLength = 10, scrollX = TRUE)) %>% 
  formatRound(c("t.stop", "x2"), 3)
```
Here, subject \#1 experienced `r sum(subset(dat.amcox, id == 1)$event)` recurrent events 
(at times `r round(subset(dat.amcox, id == 1)$t.stop[1:sum(subset(dat.amcox, id == 1)$event)], 3)`)
and died at time `r round(tail(subset(dat.amcox, id == 1)$t.stop, 1), 3)`.
Similarly, subject \#2 experienced `r sum(subset(dat.amcox, id == 2)$event)` 
recurrent events (at times `r round(subset(dat.amcox, id == 2)$t.stop[1:sum(subset(dat.amcox, id == 2)$event)], 3)`)
and is alive when censored at time `r round(tail(subset(dat.amcox, id == 2)$t.stop, 1), 3)`.

A side-by-side event plot shows the difference in the recurrent process.
```{R comp, message = FALSE}
library(ggplot2)
library(gridExtra)
grid.arrange(plotEvents(Recur(t.stop, id, event, status) ~ 1, 
                        data = dat.cox,
                        main = "type = cox") + 
             theme(legend.position = "none"),
             plotEvents(Recur(t.stop, id, event, status) ~ 1, 
                        data = dat.amcox,
                        main = "type = am|cox") + 
             theme(legend.position = "none"),
             ncol = 2)
```