---
title: "Simulating recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Simulating recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "")
```
In this vignette, we demonstrate how to use the `simGSC()` function in __`reReg`__ package
to simulate recurrent event data from a general scale-change model.
Since the scale-change model includes the Cox-type model and the
accelerated mean model as special cases, `simGSC()` can also be used to generate data from these submodels.
The `simGSC()` function allows the censoring time to be non-informative (independent given covariate)
or informative about the recurrent event process. 

## Notations
Let $N(t)$ be the number of recurrent events occurring over the interval $[0, t]$ and
$D$ be the failure time of interest subjects to right censoring by $C$.
Define the composite censoring time $Y = \min(D, C, \tau)$ and
the failure event indicator $\Delta = I\{D\le \min(C, \tau)\}$, where
$\tau$ is the maximum follow-up time.
We assume the recurrent event process $N(\cdot)$ is observed up to $Y$.
Let $X$ be a $p$-dimensional covariate vector.
Consider a random sample of $n$ subjects,
the observed data are independent and identically distributed (iid) copies of
$\{Y_i, \Delta_i, X_i, N_i(t), 0\le t\le Y_i\}$,
where the subscript $i$ denotes the index of a subject for $i= 1, \ldots, n$.
Let $m_i = N_i(Y_i)$ be the number of recurrent events the $i$th subject experienced
before time $Y_i$, then the jump times of $N_i(t)$ give the observed recurrent
event times
$t_{i1}, \ldots, t_{im_i}$ when $m_i > 0$.
Thus, the observed data can also be expressed as iid copies
of $\{Y_i, \Delta_i, X_i, m_i, (t_{i1}, \ldots, t_{m_i})\}$.
The primary interests in recurrent event data analysis often lie in
making inference about the recurrent event process and the failure event
and understanding the corresponding covariate effects.
The built-in dataset, `simDat`, is a simulated example of a recurrent event data that is generated from the
`simGSC()` function:
```{R}
library(reReg)
packageVersion("reReg")
data(simDat)
```

   
## The `simGSC` function {-}
The function `simGSC` generates the recurrent times from the joint model
\begin{equation}
  \begin{cases}
    \lambda(t) = \displaystyle Z\lambda_0(te^{X^\top\alpha})e^{X^\top\beta},&\\
    h(t) = \displaystyle Zh_0(te^{X^\top\eta})e^{X^\top\theta}, &t\in[0, \tau],
  \end{cases}
  (\#eqn:sc|sc)
\end{equation}
where $\lambda_0(t)$ is the baseline rate function for the recurrent event process,
$h_0(t)$ is the baseline hazard function for the failure time, 
$Z$ is a non-negative subject-specific latent frailty variable.
The $p$-dimensional regression coefficients
$(\alpha, \eta)$ and $(\beta, \theta)$
correspond to the shape and the size parameters, respectively.
The arguments in `simGSC` are as follow
```{r simGSC, message = FALSE}
args(simGSC)
```
The arguments are as follows

* `n` is the number of individual
* `summary` is a logical value indicating whether a brief data summary will be printed.
* `para` is a list of regression parameters. 
The list members `alpha`, `beta`, `eta`, and `theta` 
correspond to $\alpha$, $\beta$, $\eta$, and $\theta$ in \@ref(eqn:sc|sc), respectively.
* `xmat` is the $n\times p$ design matrix. 
* `censoring` is a $n$-dimensional numerical vector to specify the independent censoring time, $C$.
* `frailty` is a $n$-dimensional numerical vector to specify the frailty variable, $Z$.
* `tau` is the maximum follow-up time, $\tau$.
* `origin` is the time origin for each subject.
* `Lam0` is a single argument function used to specify the baseline cumulative rate function, $\Lambda_0(t)$.
* `Haz0` is a single argument function used to specify the baseline cumulative hazard function, $H_0(t)$.

At the default setting, the `simGSC()` function assumes $p = 2$ and the regression
parameters to be $\alpha = \eta=(0, 0)^\top$, $\beta = (-1, -1)^\top$, and $\theta = (1, 1)^\top$.
When the `xmat` and the `censoring` arguments are not specified,
the `simGSC()` function assumes $X_i$ is a two-dimensional vector $X_i = (X_{i1}, X_{i2}), i = 1, \ldots, n$,
where $X_{i1}$ is a Bernoulli variable with rate 0.5 and $X_{i2}$ is a standard normal variable.
With the default `xmat`, the censoring time $C$ is generated from
an independent uniform distribution in $[0, 2\tau X_{i1} + 2Z^2\tau(1 - X_{i1})]$.
Thus, the censoring distribution is covariate dependent and
is informative when $Z$ is not a constant.
On the other hand, when `xmat` is specified by the users,
the censoring time $C$ is generated from an independent uniform distribution $[0, 2 \tau]$.
When the `frailty` argument is not specified, the frailty variable $Z$ is generated
from a gamma distribution with a unit mean and a variance of 0.25.
The default values for `tau` and `origin` are 60 and 0, respectively.
When arguments `Lam0` and `Haz0` are left unspecified,
the `simGSC()` function uses $\Lambda_0(t) = 2\log(1 + t)$
and $H_0(t) = \log(1 + t) / 5$, respectively.
This is equivalent to setting
`Lam0 = function(x) 2 * log(1 + x)` and `Haz0 = function(x) log(1 + x) / 5`.
In summary, the default specifications generate the recurrent events and the terminal events
from the model:
$$\begin{cases}
     \lambda(t) = \displaystyle \frac{2Z}{1 + te^{-X_{i1} - X_{i2}}}, &\\
     h(t) = \displaystyle \frac{Z}{5(1 + te^{X_{i1} + X_{i2}})}, & t\in[0, 60].
   \end{cases}$$


## Examples

### `simDat`
The `simGSC()` function generates simulated data from the above specification and returns a
`data.frame` in the same format as the built-in data set, `simDat`.
Specifically, `simDat` was generated using the default settings of `simGSC()`.
```{R}
data(simDat)
set.seed(0); dat <- simGSC(200, summary = TRUE)
identical(dat, simDat)
```

### Simulating from general scale-change models
The `simDat` is an example that simulates recurrent event data when both the rate function and the hazard function 
are Cox-type model. 
```{R}
set.seed(0); datCox <- simGSC(200, summary = TRUE)
```
The hypothesis test proposed by @xu2019generalized shows a strong evidence that the rate function 
has a Cox structure.
```{R, cache = TRUE}
fit <- reReg(Recur(t.start %to% t.stop, id, event, status) ~ x1 + x2, 
             model = "gsc", B = 200, se = "sand", data = datCox)
summary(fit, test = TRUE)
```

Similarly, the hypothesis test shows a strong evidence that the rate function has
an accelerated mean or accelerated rate structure when the data is simulated under
these.

When rate has an accelerated mean structure:
```{R, cache = TRUE}
set.seed(0); datAM <- simGSC(200, para = list(alpha = c(-1, -1), beta = c(-1, -1)))
summary(update(fit, data = datAM), test = TRUE)
```

When rate has an accelerated rate structure:
```{R, cache = TRUE}
set.seed(0); datAR <- simGSC(200, para = list(alpha = c(-1, -1), beta = c(0, 0)))
summary(update(fit, data = datAR), test = TRUE)
```
 
## Reference
