---
title: "Improving Cox-type proportional rate model with `reReg`"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Improving proportional rate model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE, collapse = TRUE}
library(knitr)
opts_chunk$set(prompt = TRUE, comment = "")
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
knitr::opts_chunk$set(fig.width = 6, fig.height = 3.5) 
```

The Cox-type proportional rate model (available in `reReg()` via `model = "cox.LWYY"`)
is a common semiparametric regression model for 
recurrent event processes under the noninformative censoring assumption.
Since the construction of the pseudo-partial score function ignores the
dependency among recurrent events, 
the efficiency of the Cox-type proportional rate model can be improved by combining 
a system of weighted pseudo-partial score equations via the generalized method of moments (GMM)
and empirical likelihood (EL) estimation [@huang2022improved].
In this vignette, we demonstrate the proposed GMM and EL procedures in the `reReg` package.
We will illustrate the idea with simulated data generated from the `simGSC()` function.


```{r load-pkg, message = FALSE}
library(reReg)
packageVersion("reReg")
```

## Loading data set

We first generate a simulated data where the recurrent event process is generated from a Cox-type proportional rate model.
This can be achieved with the default settings of `simGSC()`; 
see the vignette on [simulating recurrent event data](https://www.sychiou.com/reReg/articles/reReg-sims.html) for illustrations of `simGSC()`.

```{r simDat}
set.seed(1); head(dat <- simGSC(50, frailty = rep(1, 50)))
```

## Fitting Cox-type model

We can fit the Cox-type proportional rate model [@lin2000semiparametric] with `reReg()` via the following specification. 
```{r cox, cache = TRUE}
fit <- reReg(Recur(t.stop, id, event) ~ x1 + x2, model = "cox.LWYY", data = dat)
summary(fit)
```
@huang2022improved proposed improved semiparametric estimation methods of the above proportional rate model.
By combining different weighted pseudo-partial score functions through
generalized methods of moments or empirical likelihood methods, 
@huang2022improved showed that substantial efficiency gain can be achieved without imposing additional 
model assumptions than those assumed in @lin2000semiparametric. 
Specifically, the efficiency gain is a result of combining a set of asymptotically 
independently and identically distributed estimating equations derived from the weighted pseudo-partial 
score equations of @lin2000semiparametric.

## Improved estimation

Two available approaches to combine estimating equations are the generalized method of moments (GMM)
and the empirical likelihood (EL) estimations, both of which can be specified with the `cppl` argument 
within the `control` list.
In addition to the combination approach, weight functions to be combined with 
the weighted pseudo-partial score functions can also be specified with the `cppl.wfun` argument within the `control` list.
In theory, combining more weight functions leads to better asymptotic efficiency, 
especially when the sample size is large.
However, as observed in simulation studies of @huang2022improved, 
combining one or two weighted functions can achieve substantial efficiency gain under scenarios 
with small to moderate sample sizes.
When the sample size is small, @huang2022improved suggests using a combination of the
unit weight and a decreasing function, since the decreasing function can downweight the
later time period, in which the events are more effectively censored.
For those reason, the `reReg()` currently only allows user to specify up to two weight functions with 
`cppl.wfun`.
The `reReg()` allows users choose either the cumulative baseline rate function or the Gehan's weight 
as two possible options for the weight functions.
Additionally, the `reReg()` also allows users 
to specify arbitrary function in `cppl.wfun` via function formulas.


The following gives an example of the improved estimation by combining estimating equations 
via GMM with the cumulative baseline rate function and the Gehan's weight.
```{R cox2, cache = TRUE}
fit2 <- reReg(Recur(t.stop, id, event) ~ x1 + x2, model = "cox.LWYY", data = dat, 
              control = list(cppl = "GMM", cppl.wfun = list("Gehan", "cumbase"))) 
summary(fit2) 
```
For this data set, the standard errors changed from `r fit$par1.se[1]` and `r fit$par1.se[2]` 
to `r fit2$par1.se[1]` and `r fit2$par1.se[2]`, respectively.
This achieves a `r 100 * round(1 - fit2$par1.se / fit$par1.se, 4)[1]`\% and
`r 100 * round(1 - fit2$par1.se / fit$par1.se, 4)[2]`\% efficiency gain, respectively.
The above result can be called by updating the original fit, `fit`, with the `update()` function, 
as illustrated in the following.
```{R, eval = FALSE}
fit2 <- update(fit, control = list(cppl = "GMM", cppl.wfun = list("Gehan", "cumbase")))
```


The following gives an example of the improved estimation by combining estimating equations 
via EL with the user-specified function $1 / (x + 1)$ and the Gehan's weight.
```{R cox3, cache = TRUE}
fit3 <- reReg(Recur(t.stop, id, event) ~ x1 + x2, model = "cox.LWYY", data = dat, 
              control = list(cppl = "EL", cppl.wfun = list(function(x) 1 / (x + 1), "Gehan"))) 
summary(fit3) 
```
For this data set, the standard errors changed from `r fit$par1.se[1]` and `r fit$par1.se[2]` 
to `r fit3$par1.se[1]` and `r fit3$par1.se[2]`, respectively.
This achieves a `r 100 * round(1 - fit3$par1.se / fit$par1.se, 4)[1]`\% and
`r 100 * round(1 - fit3$par1.se / fit$par1.se, 4)[2]`\% efficiency gain, respectively.
Although the above example assumes the covariates are time-independent, 
the proposed approach and the implementation allow time-dependent covariates.


## Conclusion

Although this vignette is focus on improving the proportional rate model, 
the proposed approaches by @huang2022improved can be extended to handle other types of rate function.
Such an extension will be investigated and the `reReg` package will be expanded accordingly.



## Reference
