---
title: "Fitting semiparametric regression model to recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Fitting semiparametric regression model to recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "")
```	

In this vignette, we demonstrate how to use the `reReg()` function in __`reReg`__ package
to fit semiparametric regression models to recurrent event data.

## Model assumption
A general joint model for the rate function of the recurrent event process and
the hazard function of the failure time can be formulated as follow: 
\begin{equation}
\lambda(t) = Z \lambda_0(te^{X^\top\alpha})e^{X^\top\beta};
h(t) = Z h_0(te^{X^\top\eta})e^{X^\top\theta},
\label{eq:joint}
\end{equation}
where $Z$ is a latent shared frailty variable to account for association between the two types of
outcomes, $\lambda_0(\cdot)$ is the baseline rate function,
$h_0(\cdot)$ is the baseline hazard function, and the regression coefficients
$(\alpha, \eta)$ and $(\beta, \theta)$ correspond to the shape and size parameters 
of the rate function and hazard function, respectively.
In contrast to many shared-frailty models that require a parametric assumption,
<!--so that the likelihood method can be used to estimate the joint model,-->
following the idea of @Wang2001,
the `reReg()` function implements semiparametric estimation procedures that do not require the
knowledge about the frailty distribution.
As a result, the dependence between recurrent events and failure event is left unspecified
and the proposed implementations accommodate informative censoring.
The `reReg()` function fits the recurrent event data under the above joint model setting. 
The arguments of `reReg()` are as follows

## Arguments of `reReg()`
```{r reReg-args}
library(reReg); args(reReg)
```

* `formula` a formula object, with the response on the left of a "~" operator, and the predictors on the right. 
The response must be a recurrent event survival object as returned by function `Recur`. 
See the vignette on [Visualization of recurrent event data](https://www.sychiou.com/reReg/articles/reReg-plots.html) or 
[Introduction to formula response function `Recur()`](https://wenjie-stat.me/reda/articles/reda-recur) for examples in creating `Recur` objects.
* `data` an optional data frame in which to interpret the variables occurring in the `formula`.
* `B` a numeric value specifies the number of resampling (or bootstrap) for variance estimation. 
When `B = 0`, variance estimation will not be performed.
* `method` a character string specifying the underlying model.
* `se` a character string specifying the method for standard error estimation. 
* `control` a list of control parameters.

### Choosing model type with `method`
Model \\eqref{eq:joint} includes several popular semiparametric models as special cases,
which can be specified via the `method` argument with the 
rate function and hazard function separated by "`|`".
For examples, the joint Cox model of @Huang2004 is a special case of
\\eqref{eq:joint} when $\alpha = \eta = 0$ and can be called by `method = "cox|cox"`;
the joint accelerated mean model of @xu2017joint is a special case when
$\alpha = \beta$ and $\eta = \theta$ and can be called by `method = "am|am"`.
<!--Another option is `ar` for the accerlated rate model ($\beta = \theta = 0$).-->
<!--When the primary interest is in the covariate effects on the risk of recurrent events and -->
Treating the terminal event as nuisances ($\eta = \theta = 0$),
\\eqref{eq:joint} reduces to the generalized scale-change model of
@xu2019generalized, called by `method = "sc|."`.
Moreover, users can mix the models depending on the application.
For example, `method = "cox|ar"` postulate a Cox proportional model for the recurrent event rate
function and an accelerated rate model for the terminal event hazard function
($\alpha = \theta = 0$ in \\eqref{eq:joint}). 

### Choosing the variance esitmation approach with `se`
For inference, the `reReg()` function provides several approaches for variance estimation.
The default option is `se = "resampling"`, which refers to the efficient resampling-based 
sandwich estimator.
The general idea is to decompose the limiting covariance matrix in a sandwich form, 
and its components are estimated via perturbed estimating equations.
Details of the resampling approach can be found in @zeng2008, @xu2017joint, and @xu2019generalized.
The resampling approach is faster than the conventional bootstrap, 
which can be called with `se = "bootstrap"`, 
as it only requires evaluating perturbed estimating equations rather than solving them.
When `se = "bootstrap"`, user has an option to carry out the bootstrap with parallel computing.
This can be done by specifying `parallel = TRUE` in the `control` list. 
When `parallel = TRUE`, the number of CPU cores can be specified with `parCl`.

### Control list

The complete `control` list consists of the following parameters:
  *`tol` specifies the absolute error tolerance in solving the estimating equations
  * `a0, b0` specifies the initial guesses used for root search
  * `solver` specifies the equation solver used for root search; 
  the available options are `BBsolve`, `dfsane`, `BBoptim`, and `optim` (the first three options loads the corresponding equation solver from package `BB`).
  * `baseSE` is an logical value indicating whether the confidence bounds for the baseline functions will be computed.
  * `parallel` is an logical value indicating whether parallel computation will be applied when `se = "bootstrap"` is called.
  * `parCl` is an integer value specifying the number of CPU cores to be used when `parallel = TRUE`.
  The default value is half the CPU cores on the current host.


## Examples
We will illustrate the usage of `reReg` with simulated data generated from `simSC`. 
Readers are referred to the vignette on [Simulating recurrent event data](https://www.sychiou.com/reReg/articles/reReg-sims.html)
for using `simSC` to generate recurrent event data.

### Joint Cox model of @Huang2004
A simulated model following the joint Cox model of @Huang2004 can be generated by
```{R, cox, cache = TRUE}
set.seed(1); datCox <- simSC(500, c(1, -1), c(-1, 1), summary = TRUE)
```
The underlying true model has the form:
$$\lambda(t) = Z \lambda_0(t)e^{X_1 - X_2};
h(t) = Z h_0(t)e^{-X_1 + X_2}.$$
The model fit is
```{R, coxfit, cache = TRUE}
fit.cox <- reReg(Recur(Time, id, event, status) ~ x1 + x2, data = datCox)
summary(fit.cox)
```
The baseline functions can be plotted via `plot()`: 
```{R, coxplot}
plot(fit.cox)
```

### Joint accelerated mean model of @xu2017joint
A simulated model following the joint accelerated mean model of @xu2017joint can be generated by
```{R, am, cache = TRUE}
set.seed(1); datAM <- simSC(500, c(1, -1), c(-1, 1), type = "am", summary = TRUE)
```
The underlying true model has the form:
$$\lambda(t) = Z \lambda_0(te^{X_1 - X_2})e^{X_1 - X_2};
h(t) = Z h_0(te^{-X_1 + X_2})e^{-X_1 + X_2}.$$
The model fit is
```{R, amfit, cache = TRUE}
fit.am <- reReg(Recur(Time, id, event, status) ~ x1 + x2, data = datAM, method = "am")
summary(fit.am)
```
The baseline functions can be plotted via `plot()`: 
```{R, amplot}
plot(fit.am)
```

### @xu2019generalized `sc.XCYH` 
A simulated model following the generalized scale-change model of @xu2019generalized can be generated by
```{R, sc, cache = TRUE}
set.seed(1); datSC <- simSC(500, a1 = c(1, -1), a2 = c(-1, 1), b1 = c(1, -1), b2 = c(-1, 1), type = "sc", summary = TRUE)
```
The underlying true model has the form:
$$\lambda(t) = Z \lambda_0(te^{X_1 - X_2})e^{-X_1 + X_2};
h(t) = Z h_0(te^{X_1 - X_2})e^{-X_1 + X_2}.$$
The model fit is
```{R, scfit, cache = TRUE}
fit.sc <- reReg(Recur(Time, id, event, status) ~ x1 + x2, data = datSC, method = "sc")
summary(fit.sc)
```
The baseline functions can be plotted via `plot()`: 
```{R, scplot}
plot(fit.sc)
```

### Joint Cox/accelerated rate model
A simulated model following the generalized scale-change model of @xu2019generalized can be generated by
```{R, coxar, cache = TRUE}
set.seed(1); datCoxar <- simSC(500, c(1, -1), c(-1, 1), type = "cox|ar", summary = TRUE)
```
The underlying true model has the form:
$$\lambda(t) = Z \lambda_0(t)e^{X_1 - X_2};
h(t) = Z h_0(te^{-X_1 + X_2}).$$
The model fit is
```{R, coxarfit, cache = TRUE}
fit.coxar <- reReg(Recur(Time, id, event, status) ~ x1 + x2, data = datCoxar, method = "cox|ar")
summary(fit.coxar)
```
The baseline functions can be plotted via `plot()`: 
```{R, coxarplot}
plot(fit.coxar)
```

## Bootstrap
Parallel computing is possible via specifying the control list.
```{r bootstrap, cache = TRUE}
system.time(fitB1 <- reReg(Recur(Time, id, event, status) ~ x1 + x2, data = datCox, 
                           method = "cox", se = "bootstrap"))
summary(fitB1)
system.time(fitB2 <- reReg(Recur(Time, id, event, status) ~ x1 + x2, data = datCox, 
                           method = "cox", se = "bootstrap", 
                           control = list(parallel = TRUE, parCl = 8)))
summary(fitB2)
```

## Other popular methods
Some methods that assumes \code{Z = 1} and requires independent
censorings are also implemented in \code{reReg}. 
These includes the methods proposed by 
@lin2000semiparametric, @ghosh2002, and @ghosh2003, that can be called by specifying 
`method = "cox.LWYY"`, `method = "cox.GL"`, and `method = "am.GL"`, respectively.

## Reference