---
title: "Fitting semiparametric regression model to recurrent event data"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/reReg.bib
vignette: >
  %\VignetteIndexEntry{Fitting semiparametric regression model to recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "")
```	

In this vignette, we demonstrate how to use the `reReg` function in __`reReg`__ package
to fit semiparametric regression model to recurrent event data.

```{r reReg-args}
library(reReg)
args(reReg)
```
The arguments are as follows


* `formula` a formula object, with the response on the left of a "~" operator, and the predictors on the right. The response must be a recurrent event survival object as returned by function `reSurv`. See the vignette on [Visualization of recurrent event data](https://www.sychiou.com/reReg/articles/reReg-plots.html) for examples in creating `reSurv` objects.
* `data` an optional data frame in which to interpret the variables occurring in the `formula`.
* `B` a numeric value specifies the number of resampling for variance estimation. When `B = 0`, variance estimation will not be performed.
* `method` a character string specifying the underlying model.
* `se` a character string specifying the method for standard error estimation. 
* `contrasts` an optional list.
* `control` a list of control parameters.

We will illustrate the usage of `reReg` with simulated data generated from `simSC`. 
Readers are referred to the vignette on [Simulating recurrent event data](https://www.sychiou.com/reReg/articles/reReg-sims.html)
for using `simSC` to generate recurrent event data.

## Point estimation
We give brief description on each of the available methods below.

### `cox.LWYY`

```{r cox-lwyy}
set.seed(123)
dat.cox <- simSC(n = 500, a = c(1, -1), b = c(1, -1))
fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox)
summary(fit)
```
### `cox.GL`

```{r cox-GL, tidy = TRUE}
fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox, 
             method = "cox.GL")
summary(fit)
```

### `cox.HW`
```{r cox-HW, tidy = TRUE}
fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox, 
             method = "cox.HW")
summary(fit)
```

### `am.GL`
```{r am-GL, tidy = TRUE}
set.seed(123)
dat.am <- simSC(n = 500, a = c(1, -1), b = c(1, -1), type = "am")
fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.am, 
             method = "am.GL")
summary(fit)
```

### `am.XCHWY`
```{r am-XCHWY, tidy = TRUE}
fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.am, 
             method = "am.XCHWY")
summary(fit)
```

### `sc.XCYH`
Fitting recurrent event data generated from the Cox model:
```{r sc-XCYH-cox, tidy = TRUE}
fit1 <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox, 
              method = "sc.XCYH")
summary(fit1)
```
Fitting recurrent event data generated from the accelerated model:
```{r sc-XCYH-am, tidy = TRUE}
fit2 <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox, 
              method = "sc.XCYH")
summary(fit2)
```

## Variance estimation
The variance estimation was not performed in the examples above.
However, the package provides several approaches for variance estimation.
We will take `cox.HW` for example.

### Bootstrap
```{r bootstrap, tidy = TRUE}
system.time(fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox, 
                         method = "cox.HW", se = "bootstrap"))
summary(fit)
```

### Bootstrap with parallel computing
Detect the number of CPU cores on the current host.
```{r bootstrap-detect}
parallel::detectCores()
```
The default number of CPU cores used in the parallel computing is half of the available CPU cores.
We will use all the CPU cores available.
```{r bootstrap-para, cache = TRUE, tidy = TRUE}
system.time(fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox, 
                         method = "cox.HW", se = "bootstrap", 
                         control = list(parallel = TRUE, parCl = 8)))
summary(fit)
```

### Sandwich (resampling) variance estimator

```{r bootstrap-resampling, tidy = TRUE}
system.time(fit <- reReg(reSurv(Time, id, event, status) ~ x1 + x2, data = dat.cox, 
                         method = "cox.HW", se = "resampling"))
summary(fit)
```